---
title: "Heart Prediction Model"
author: "Farrah Johnson and Mariam Ikram"
date: "2026-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Intro: 
Heart disease is one of the leading causes of death worldwide, and the number 1 cause of death in the United States, about ... Americans from all ... (should i go into stats?) (citation) Understanding the its risk factors in pivotal for medical professionals and the general public ...

Data Exploration: (what question we're answering and exploring, background on the data: where we got it from, the link, the author, the reasoning for the data being made available, important variables and their meanings)

Packages: 
tidyverse: data wrangling (dplyr), plotting (ggplot2) and general data manipulation 
ggplot2
dyplr
janitor #cleans columns names and simple tabulations

```{r, message=FALSE, warning=FALSE, include=FALSE}
# tidyverse: data wrangling (dplyr), plotting (ggplot2) and general data manipulation
library(tidyverse)
library(ggplot2) #remove this
library(dplyr) #remove this
# janitor: cleans columns names and simple tabulations
library(janitor)
```

```{r}
#Import the CSV and clean column names for easier coding
heart_raw <- read.csv("heart.csv")

heart <- heart_raw %>% 
  clean_names() # converts column names to snake_case, lowercase, no spaces

#Inspect structure after cleaning
glimpse(heart)
```
We first read in the CSV file and use `clean_names()` to standardize the column names. This makes it easier to reference variables in the rest of the analysis (for example, `chest_pain_type` instead of `ChestPainType`).


```{r, include = FALSE}

# takes you to the full database without including it in the knit

view(heart)
```

```{r}
names(heart)
range(heart$age)
```




```{r}

# Pulls out necessary data and changes the heart disease to a string to make it more decipherable 


set.seed(123)
random_sample <- heart %>%
  sample_n(20)

#I set a seed here so the data we're pulling is more random, as head(20) will only pull the first 20. 

random_sample %>%
  select(age, sex, heart_disease, resting_bp, cholesterol, max_hr) %>%
  mutate(heart_disease = recode(heart_disease,
                                "1" = "Yes",
                                "0" = "No")) %>%
  arrange(age) %>%
  head(20)

# gives an average for necessary data for a better comparison


heart %>%
  select(age, sex, heart_disease, resting_bp, cholesterol, max_hr) %>%
  mutate(heart_disease = recode(heart_disease,
                                "1" = "Yes",
                                "0" = "No")) %>%
  group_by(heart_disease) %>%
  summarise("Average Blood Pressure" = mean(resting_bp),
            "Average age" = mean(age),
            "cholesterol" = mean(cholesterol),
            "Max Heart Rate" = mean(max_hr))






```

```{r}

# cleaned heart data so the graph shows no zero values, histogram w/ facet wrap, changed the labels so it was more decipherable. I used cholesterol for this data because it seemed to have the highest correlation with heart disease, but I'll have to look into it further.

heart_clean <- heart %>%
  filter(cholesterol != 0)


ggplot(heart_clean, aes(x = cholesterol, fill = "pink")) + geom_histogram(binwidth = 20) + facet_wrap(~heart_disease, labeller = labeller(heart_disease = c("1" = "Has Heart Disease", "0" = "No Heart Disease"))) + labs(title = "cholesterol distribution",
                                                                                                                                                                                                                        x = "cholesterol level",
                                                                                                                                                                                                                        y = "People"
)


```






























